---
title: "R Basics"
output: html_document

---

#### Layout of RStudio

First off let's get acquainted with RStudio. Open it up and let's poke around a bit. 

<br>

#### Downloading R workshop folder from Github

Ideally, everyone should have already downloaded this folder. But let's go through how to do it now. 

<br>

#### File paths

File paths are the location of files on your computer. Anytime you load a file into R or save a file you created in R you need to provide a file path. These paths are different between mac and windows. 

Windows look like this : `C:/Users/G02.047/Documents/GitHub/R_Workshop/Basic_data_manipulation/colony_size.txt`  

Mac paths look like this: `/Users/tessbrewer/Documents/munich/R_Workshop/Basic_data_manipulation/colony_size.txt`

These are both absolute paths. You can make your life easier by tab completing these paths. 

<br>

#### Setting working directories and relative paths 

You can also use relative paths by setting a working directory. Here you tell R that all files you will be using are contained within this folder:
```{R, setwd}
setwd('/Users/tessbrewer/Documents/munich/R_Workshop/')
```

After this command you can use relative file paths, so something like this: `Basic_data_manipulation/colony_size.txt`  

Instead of this: `/Users/tessbrewer/Documents/munich/R_Workshop/Basic_graphing/colony_size.txt`

<br>

#### Getting help

If you're not sure how a particular command works in R, your first step is always looking at the help page. For example, if we have questions about `read.delim`, the first thing we can do is  


```{r, questions}
? read.delim
```

This brings up the R documentation for this specific command.  

If this page doesn't answer your question I highly recommend googling your issue. Many people have struggled with many things and some of them were nice enough to post online about their solutions. 

<br>

#### Loading and installing packages
R packages are collections of functions and tools developed by people who use R. They can be very helpful, and some (like ggplot2) are practically essential. Here are some commands related to their installation and loading.

  * `install.packages`: installs package
  * `library`: loads installed package. You must load the packages you'll use each time you restart a session of R. 

<br>

#### Basic syntax and operators

In R you can assign a variable in two different ways: 

  * `y = 4`
  * `x <- 4`
  
There is no difference between the two. 

We'll also go over some operators, but just the ones that will be useful for our exercises.  

The following operators will return a boolean value (TRUE or FALSE):

  * `a == b`: tests for equality
  * `a != b`: tests for inequality (generally a "!" in R negates something)
  * `a > b`:  tests for greater than
  * `a >= b`: tests for greater than or equal to
  * `%in%`: tests whether an element belongs to a vector
  
Lastly we have:

  * `$` lets you access objects stored within other objects

<br>

#### Types of data

In R we have two general types of data: continuous and categorical. 

  * Continuous data is numeric and has a natural order.
    + example = `c(1, 3, 6, 9)`
  * Categorical data does not necessarily have a natural order and are based on finite groups. 
    +  example = `c('hamburger', 'hot dog', 'schnitzel')`


<br>

## Loading data and accessing elements of a dataframe

We'll start with reading a file into R.  

**Change this path to where you saved the R_Workshop folder.**


```{r, reading_data, message=FALSE, strip.white=TRUE, fig.height=5.5, fig.width=12, fig.align='center'}

colony_size <- read.delim('~/Documents/munich/R_Workshop/Basic_data_manipulation/colony_size_updated.txt')
class(colony_size) 

```

`read.delim` returns a data.frame object. Let's take a look at it.  

```{r, reading_data_head, message=FALSE}

head(colony_size)

```

This is colony size data exported from the Analyze Particles function in ImageJ. Every row is data for one colony, and different plates had different numbers of colonies.  

We can see we have **categorical data** (pbad24, pbad33) and **continuous data** (area, width, height).

This is a complementation assay. A ΔefpΔuup strain was complemented with one of nine different EF-P carried on pBAD24. 

We can check the species each EF-P came from using `table`. This is an R function I use constantly, it returns a table with counts data.

```{r, table, message=FALSE}
table(colony_size$efp_species)
```

The effect of expression of EarP on pBAD33 was then tested to see if this impacts growth rescue. 

Let's play with the dataframe.  

<br>

#### Columns

The basic layout of dataframes is `df[rows, columns]`.

You can access columns in several ways. By coordinates:

```{r, accessing dataframe_coor}
head(colony_size[, 1]) 
```

By column names:
```{r, accessing dataframe_colname}
head(colony_size[, 'Area']) 
```

Or by column names with the dollar sign:
```{r, accessing dataframe_dollar}
head(colony_size$Area) 
```
This is my favorite, it's most human readable in my opinion. 

These methods are all equivalent. But you don't need to take my word for it! We can double check:
```{r, fgfd}
table(colony_size[, 'Area'] == colony_size[, 1])
```

Why does it say TRUE 1442? Because R is comparing each row in the column "Area", and there are `r nrow(colony_size)` rows, each of which returned TRUE for the comparison.

<br>

#### Rows

Rows can be accessed this way:

```{r, rows}
colony_size[1, ] 
```

<br>

#### Cells

And specific cells of the dataframe can be accessed like this:

```{r, cells}
colony_size[1, 1]
```
<br>

## Subsetting your dataframe

Oftentimes you want to get a subset of your dataframe. Let's see which different variables we have and then we'll subset for one in particular.

```{r, tables}
table(colony_size$genotype)
```

Let's say we want to subset our dataframe to include ONLY rows that are from the genotype 'deltaefpuup_deac24empty33'. 


We can do this by using the coordinate system:

```{r, subset with coordinates}
check_coor <- colony_size[colony_size$genotype == 'deltaefpuup_deac24empty33',]
table(check_coor$genotype)
```

Or we can use `subset`, which is a base r function -- it's very human readable.

```{r, subset with subset whole}
check_subset <- subset(colony_size, genotype == 'deltaefpuup_deac24empty33')
table(check_subset$genotype)
```

<br>

## Let R do the dirty work

Say you want to subset the dataframe for only the three samples with the most colonies.  

You can first use `table` to see which samples have the most entries, then manually select for them by typing.  

We can do that like this:

```{r, hard code}
sort(table(colony_size$genotype))

check_3_hardcoded <- subset(colony_size, genotype %in% 
                     c('deltaefpuup_heau24earp33', 'deltaefpuup_eco24earp33', 'deltaefpuup_ppu24earp33'))

table(check_3_hardcoded$genotype)
```

This is called hard-coding data and usually not the best way to go about things in R. Say you want to use this same script for another set of colony sizes. You'd have to rewrite this whole part!   

Instead, we can make R do this for us.

```{r, non-hard coded}
tail(sort(table(colony_size$genotype)), n = 3) #those are the three we want

names(tail(sort(table(colony_size$genotype)), n = 3)) #get just their names, not their frequency counts

what_we_want <- names(tail(sort(table(colony_size$genotype)), n = 3)) #write those to a variable

check_3_lazy <- subset(colony_size, genotype %in% what_we_want) #use that to subset

table(check_3_lazy$genotype)
```

<br>

## Modifying dataframes

Oftentimes we want to modify dataframes. For instance, maybe you want to rename a column name:
```{r, modifying dataframes}

colnames(check_subset) #list the column names of our dataframe

colnames(check_subset)[1] #check first column name
colnames(check_subset)[1] <- 'butts'
colnames(check_subset) #success

```

<br>

We can also add new columns. You can add them very simply, like this:

```{r, new columns}
check_subset$three <- 3
head(check_subset$three)
```

We can remove columns too.

```{r, remove columns}
check_subset$three <- NULL #by column name
check_subset <- check_subset[, -3] #remove third column 
```



We can also use `ifelse` statements to add columns. Either rather simply:
```{r ifelse simple}
colony_size$max_dimension <- ifelse(colony_size$Height > colony_size$Width, 
                                    colony_size$Height, 
                                    colony_size$Width)
head(colony_size[, c(2:3, 8)])
```

Or slightly more complex columns that are dependent on other values. For example:
```{r, ifelse}
colony_size$type <- ifelse(grepl('deltaefpuup_empty24', colony_size$genotype) == TRUE, 
                           'control', 
                           'experimental')
table(colony_size$genotype, colony_size$type)
```
<br>

## Saving your files
We added two new columns to our dataframe that we might want to use later on. So, let's save this new dataframe with `write.table`!  

#### Delimiters
I usually save my files in the `.txt` format. These type of files use TAB characters for delimiters.  

Delimiters basically tell the computer where the boundary between columns is. For example, `.csv` files use commas as delimiters.  

Save it however you want, but you must tell R what you want as the delimiter.  

In the case of `.txt` files, that's  `sep = '\t'`. For `.csv` files it would be `sep = ','`

<br>

#### Row.names in R

`row.names` are special rownames you can provide for a dataframe. They have strict requirements:

  1. Must be exact length of the number of rows, with NO missing values
  2. Cannot contain duplicate names

They are useful in certain situations (you can ask me if you're interested).  

In these intro courses however, we don't use them, and for that reason I put `row.names = FALSE`.

<br>

#### Don't make R mad. 

There are some general guidelines for naming files you will access in R.  

Do not use any special characters in file names, or spaces.  

Do not start column names with numbers. R does not like this!

Otherwise, try to use common sense. Don't use commas within columns if your file is `.csv`. This is asking for trouble!


```{r, saving}
? write.table

write.table(colony_size, '~/Documents/munich/R_Workshop/Basic_data_manipulation/colony_size_saved.txt', 
            row.names = FALSE, sep = '\t')
```